\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2020-08-18 analysis/paper/paper.Rmd
time dependent hazard ratio for ERSPC

options(kableExtra.latex.load_packages = FALSE)
library(casebase)
library(colorspace)
library(cowplot)
library(data.table)
library(glmnet)
library(kableExtra)
library(magrittr)
library(pracma)
library(flexsurv) # for fitting gompertz AFT
library(prodlim)
library(riskRegression)
library(splines)
library(survival)
library(tibble)
library(visreg)
library(lubridate)
library(tidyverse)
library(scales)
options(digits = 2, scipen = 999)
# set the seed for reproducible output
set.seed(1234)

@@

The case-base sampling approach described in Section \ref{theory} can be visualized in the form of a population time plot. These plots are extremely informative graphical displays of survival data and should be one of the first steps in an exploratory data analysis. The \code{popTime} function and \code{plot} method facilitate this task:

@@

To illustrate these functions, we will use the European Randomized Study of Prostate Cancer Screening (ERSPC) data [@schroder2009screening] which was extracted using the approach described in Liu *et al.* [-@liu2014recovering] This dataset is available through the \pkg{casebase} package. It contains the individual observations for 159,893 men  from seven European countries, who were between the ages of 55 and 69 years when recruited for the trial.

@@

```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plot for ERSPC data."}
label_number_si <- function (accuracy = 1, unit = NULL, sep = NULL, ...) {
    sep <- if (is.null(unit)) 
        ""
    else " "
    scales:::force_all(accuracy, ...)
    function(x) {
        breaks <- c(0, 10^c(K = 3, M = 6, B = 9, T = 12))
        n_suffix <- cut(abs(x), breaks = c(unname(breaks), Inf), 
            labels = c(names(breaks)), right = FALSE)
        n_suffix[is.na(n_suffix)] <- ""
        suffix <- paste0(sep, n_suffix, unit)
        scale <- 1/breaks[n_suffix]
        scale[which(scale %in% c(Inf, NA))] <- 1
        number(x, accuracy = accuracy, scale = unname(scale), 
            suffix = suffix, ...)
    }
}
plot(pt_object, 
     add.base.series = TRUE,
     ratio = 0.7,
     facet.params = list(ncol = 2),
     ribbon.params = list(fill = "gray50"),
     case.params = list(size = 0.8),
     base.params = list(size = 0.8)) + 
  paper_gg_theme +  
  scale_y_continuous(labels = label_number_si())
```

@@

Next, layers for the case series and base series are added. The case series is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area. By randomly distributing the cases, we can get a better sense of the incidence density. In Figure \ref{fig:plot-stratified-erspc-data}, we see that more events are observed at later follow-up times. Therefore, a constant hazard model would not be appropriate in this instance as it would overestimate the cumulative incidence earlier on in time, and underestimate it later on. Finally, the base series is sampled horizontally with sampling weight proportional to their follow-up time. The reader should refer to the package vignettes for more examples and a detailed description of how to modify the aesthetics of a population-time plot.

@@

For our first case study, we return to the ERSPC study and investigate the differences in survival between the control and screening arms. We fit four models that differ in which functional form of time is used: 1) excluded from the linear predictor, 2) linear function, 3) log function, and 4) a smooth function using cubic B-splines. The models are fit using \code{fitSmoothHazard}. Since the output object from \code{fitSmoothHazard} inherits from the \code{glm} class, we can directly use the function \code{summary}:

@@

```{r, eval = eval_cs1, echo=FALSE}
summary(fits[["gompertz"]])
```

@@

>>>>

options(kableExtra.latex.load_packages = FALSE)
library(casebase)
library(colorspace)
library(cowplot)
library(data.table)
library(glmnet)
library(kableExtra)
library(magrittr)
library(pracma)
library(flexsurv) # for fitting gompertz AFT
library(prodlim)
library(riskRegression)
library(splines)
library(survival)
library(tibble)
library(visreg)
library(lubridate)
library(tidyverse)
library(scales)
options(digits = 2)
# set the seed for reproducible output
set.seed(1234)

@@

Population-time plots are a descriptive visualization of incidence density, where aggregate person-time is represented by area and events as points within the area. The case-base sampling approach described in Section \ref{theory} can be visualized in the form of a population time plot. These plots are extremely informative graphical displays of survival data and should be one of the first steps in an exploratory data analysis. The \code{popTime} function and \code{plot} method facilitate this task:

@@

To illustrate these functions, we will use data from the European Randomized Study of Prostate Cancer Screening (ERSPC) [@schroder2009screening] which was extracted using the approach described in Liu *et al.* [-@liu2014recovering] This dataset is available through the \pkg{casebase} package. It contains the individual observations for 159,893 men  from seven European countries, who were between the ages of 55 and 69 years when recruited for the trial.

@@

```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plots for both treatment arms in the ERSPC dataset. The grey area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely small rectangles (person-moments). The points shown in the figure are select person-moments from the case and base series. We see that more events are observed at later follow-up times, motivating the use of a non-constant in time hazard model."}
label_number_si <- function (accuracy = 1, unit = NULL, sep = NULL, ...) {
    sep <- if (is.null(unit)) 
        ""
    else " "
    scales:::force_all(accuracy, ...)
    function(x) {
        breaks <- c(0, 10^c(K = 3, M = 6, B = 9, T = 12))
        n_suffix <- cut(abs(x), breaks = c(unname(breaks), Inf), 
            labels = c(names(breaks)), right = FALSE)
        n_suffix[is.na(n_suffix)] <- ""
        suffix <- paste0(sep, n_suffix, unit)
        scale <- 1/breaks[n_suffix]
        scale[which(scale %in% c(Inf, NA))] <- 1
        number(x, accuracy = accuracy, scale = unname(scale), 
            suffix = suffix, ...)
    }
}
plot(pt_object, 
     add.base.series = TRUE,
     ratio = 0.7,
     facet.params = list(ncol = 2),
     ribbon.params = list(fill = "gray50"),
     case.params = list(size = 0.8),
     base.params = list(size = 0.8)) + 
  paper_gg_theme +  
  scale_y_continuous(labels = label_number_si()) + 
  ylab("Population Size")
```

@@

Next, layers for the case series and base series are added. The y-axis location of each case moment is sampled at random vertically on the plot to avoid having all points along the upper edge of the gray area. By randomly distributing the cases, we can get a sense of the incidence density. In Figure \ref{fig:plot-stratified-erspc-data}, we see that more events are observed at later follow-up times. Therefore, a constant hazard model would not be appropriate in this instance as it would overestimate the cumulative incidence earlier on in time, and underestimate it later on. Finally, the base series is sampled horizontally with sampling weight proportional to their follow-up time. The reader should refer to the package vignettes for more examples and a detailed description of how to modify the aesthetics of a population-time plot.

@@

For our first case study, we return to the ERSPC study and investigate the differences in risk between the control and screening arms. We fit four models that differ in which functional form of time is used: 1) excluded from the linear predictor, 2) linear function, 3) log function, and 4) a smooth function using cubic B-splines. The models are fit using \code{fitSmoothHazard}. Since the output object from \code{fitSmoothHazard} inherits from the \code{glm} class, we can directly use the function \code{summary}:

@@

```{r, eval = eval_cs1, echo=FALSE}
print.summary.glm <- function(x,digits=max(3,getOption("digits")-3),symbolic.cor=x$symbolic.cor,
                              signif.stars = getOption("show.signif.stars"),show.call=FALSE,show.resids=FALSE,...) {
  if (show.call) cat("Call:\n", paste(deparse(x$call), sep = "\n", collapse = "\n"), "\n", sep = "")
  if (show.resids) {
    cat("Deviance Residuals: \n")
    if (x$df.residual > 5) {
      x$deviance.resid <- stats::quantile(x$deviance.resid, na.rm = TRUE)
      names(x$deviance.resid) <- c("Min", "1Q", "Median", "3Q", "Max")
    }
    xx <- zapsmall(x$deviance.resid, digits + 1)
    print.default(xx, digits = digits, na.print = "", print.gap = 2)
    cat("\n")
  }
  if (length(x$aliased) == 0L) cat("No Coefficients\n")
  else {
    df <- if ("df" %in% names(x)) x[["df"]]
    else NULL
    if (!is.null(df) && (nsingular <- df[3L] - df[1L])) 
      cat("Coefficients: (", nsingular, " not defined because of singularities)\n", sep = "")
    else cat("\nCoefficients:\n")
    coefs <- x$coefficients
    if (!is.null(aliased <- x$aliased) && any(aliased)) {
      cn <- names(aliased)
      coefs <- matrix(NA, length(aliased), 4L, dimnames = list(cn,colnames(coefs)))
      coefs[!aliased, ] <- x$coefficients
    }
    stats::printCoefmat(coefs, digits = digits, signif.stars = signif.stars,na.print = "NA", ...)
  }
  cat("\n(Dispersion parameter for ", x$family$family, " family taken to be ", 
      format(x$dispersion), ")\n\n", apply(cbind(paste(format(c("Null","Residual"), 
                                                              justify = "right"), "deviance:"),
                                                 format(unlist(x[c("null.deviance","deviance")]), digits = max(5, digits + 1)), " on", 
                                                 format(unlist(x[c("df.null", "df.residual")])), " degrees of freedom\n"),1L,
                                           paste, collapse = " "), sep = "")
  if (nzchar(mess <- stats::naprint(x$na.action))) cat("  (", mess, ")\n",sep="")
  cat("AIC: ",format(x$aic,digits=max(4,digits+1)),"\n\n",
      "Number of Fisher Scoring iterations: ",x$iter,"\n",sep="")
  correl <- x$correlation
  if (!is.null(correl)) {
    p <- NCOL(correl)
    if (p > 1) {
      cat("\nCorrelation of Coefficients:\n")
      if (is.logical(symbolic.cor) && symbolic.cor) print(stats::symnum(correl,abbr.colnames=NULL))
      else {
        correl <- format(round(correl, 2),nsmall=2,digits=digits)
        correl[!lower.tri(correl)] <- ""
        print(correl[-1,-p,drop=FALSE],quote=FALSE)
      }
    }
  }
  cat("\n")
  invisible(x)
}

print(summary(fits[["gompertz"]]), signif.stars = F)
```

@@

```{r interaction-ERSPC, eval = eval_cs1}
fit_inter <- casebase::fitSmoothHazard(DeadOfPrCa ~ poly(Follow.Up.Time, 2)*ScrArm, data = ERSPC)
summary(fit_inter)

# reference category
newdata <- data.frame(ScrArm = factor("Control group",
                                         levels = c("Control group","Screening group")),
                      Follow.Up.Time = seq(1,12,by =0.1))
str(newdata)

plot(fit_inter,
     type = "hr",
     newdata = newdata,
     var = "ScrArm",
     increment = 1,
     xvar = "Follow.Up.Time",
     ci = T,
     rug = T)
abline(a=1,b=0, lty = 2)
```
